---
title: "Practical E"
author: "Anne Vinkel, adapted from Mark van der Loo "
date: "Statistical Programming with R"
output: html_document
---

### 1 - a bit on factors
**Define the vector**
```{r}
v <- factor(c(1,8.2,3,11))
```
**What happens if you apply the function `as.numeric` to `v`? How can you make `v` into a numeric vector?**

```{r}
as.numeric(v)
```
Since `v` is a factor, the visible values in it are actually labels for the underlying factor values - which are nicely ordered consequtive integers. In order to make `v` numeric with the *right* values, use
```{r}
as.numeric(as.character(v))
```


### 2 - reading in data with base R and `readr`

**The file `unnamed.csv` contains comma-separated values of ages and heights for a set of imaginary persons. Read in the data. Make sure to name both variables, and make both numeric. The obvious candidate function is `read.table`, but the `read_csv` function in the `readr` package makes it a lot easier** 

Let's try with `read.table`
```{r}
## reading with read.table ----
dat <- read.table(
  file = "data/unnamed.csv"
  , header = FALSE
  , col.names = c("age","height")
  , stringsAsFactors=FALSE
  , sep=",")
dat
```
The file reads in nicely, and we get useful variable names, but the data in the `height` variable isn't clean and cannot be read as numeric. We can try to coerce the variable into numeric, but this fails:

```{r, error = T}
## define colClasses ----
dat <- read.table(
  file = "data/unnamed.csv"
  , header = FALSE
  , col.names = c("age","height")
  , colClasses = c("numeric","numeric")
  , stringsAsFactors=FALSE
  , sep=",")
```
Instead, we can read as character, clean up the text and convert to numeric. Or we could go the tidyverse way or use the `readcsv` function in the `readr` package. 

```{r, error = T}
# parse columns as 'number' (flexible)
readr::read_csv("data/unnamed.csv"
      , col_names=c("age","height")
      , col_types="nn")

```

### 3 - matching dirty data against standard labels
**The file `bands.txt` has a list of bands, but somebody messed up the "country" variable. Match it against the vector `c("Australia", "United States", "United Kingdom", "Germany", "Spain")`. Also make sure the `date` variable is a date.**

We read in the data nicely with the `readr` package:
```{r}
library(readr)
bands <- read_delim("data/bands.txt", delim = "\t", trim_ws = T,
                     col_types = cols( date = col_date(format = "%d%b%Y")) )
```

Now for the label matching
```{r}
countries <- trimws(bands$country)
codes <- c("Australia", "United States", "United Kingdom", "Germany", "Spain")

## base R match function gives the clean matches
match(countries, codes)

## Which values are failing? 
countries[is.na(match(countries, codes))]

## The amatch function allows for some slack
library(stringdist)
amatch(countries, codes, maxDist=3)

## Which values are still failing?
countries[is.na(amatch(countries, codes, maxDist=3))]

## We match the closest fit
D <- stringdistmatrix(countries, codes)
i <- apply(D, 1, which.min)
data.frame(rawtext = countries, coded = codes[i])

## This may be the time when we hand-code the rest:
bands$country_clean <- codes[amatch(countries, codes, maxDist=3)]
bands$country_clean[ bands$country == "UK"] <- "United Kingdom"
bands$country_clean[ bands$country == "US"] <- "United States"
```


