---
title: "Practical E"
author: "Anne Vinkel, adapted from 'An Introduction to data cleaning with R' by Edwin de Jonge and  Mark van der Loo "
date: "Statistical Programming with R"
output: html_document
---

### 1 - a bit on factors
**Define the vector**
```{r}
v <- factor(c(1,8.2,3,11))
```
**What happens if you apply the function `as.numeric` to `v`? How can you make `v` into a numeric vector?**

---

### 2 - reading in data with base R and `readr`

**The file `unnamed.csv` contains comma-separated values of ages and heights for a set of imaginary persons. Read in the data. Make sure to name both variables, and make both numeric. The obvious candidate function is `read.table`, but the `read_csv` function in the `readr` package makes it a lot easier** 

---

### 3a - data cleaning
**Exercise 3 is about removing inconsistencies from data. The exercise comes in two versions: One using base procedures by "hand", and one using the editrules package. This first version is less abstract and does not require looking into a new framework. Feel free to skip it and do  3b instead, if it is too simple - or feel free to skip 3b if it's too involved**

**We will work with the following data - you can find it on the course homepage or paste it into a text file from here:**
```{r, eval=F}
age, agegroup, height, status, yearsmarried
21, adult, 6.0, single, -1
2, child, 3, married, 0
18, adult, 5.7, married, 20
221, elderly, 5, widowed, 2
34, child, -7, married, 3
```

```{r}
people <- read.csv( file = "data/people.txt", stringsAsFactors = F, strip.white = T)
```
**Correct the data, replacing impossible values with correct ones or with `NA` where the correct value cannot be deduced:**

* **Age should not be negative and should be less than 150**
* **Height should be greater than zero**
* **Age should be greater than the number of years married**
* **Children cannot be married or widowed**
* **People under 18 of age should be coded as children, those between 18 and 64 should be coded as adults, and those over age 65 should be coded as elderly**
* **Age should be larger than or equal to yearsmarried + 18**

--- 

### 3b - a walkthrough for the `editrules` package 

**This exercise is adapted from an example in De Jonge and van der Loo and walks you through applying functions in the `editrules` package. The `editrules` package allows allows us to define rules which each record of a data set must obey. Editrules can check which rules fail for each record, and find the minimal set of variables to adapt so that all rules can be folowed.** 

**We will work with the following data - you can find it on the course homepage or paste it into a text file from here:**
```{r, eval=F}
age, agegroup, height, status, yearsmarried
21, adult, 6.0, single, -1
2, child, 3, married, 0
18, adult, 5.7, married, 20
221, elderly, 5, widowed, 2
34, child, -7, married, 3
```

```{r}
people <- read.csv( file = "data/people.txt", stringsAsFactors = F, strip.white = T)
```
**For a cleaner workflow, the rules that these data must obey are treated as a separate object, that can then be stored and reapplied (instead, say, of treating these rules and checks as lines of code in a syntax file). We define an edit set and confront the data with these rules:**
```{r, warning=F, message=F}
library(editrules)
E <- editset( c( "age >=0" , "age <= 150" ) )
E
violatedEdits( E, people)
```
**A realistic set of edit rules will be rather larger. It may be defined in a text file and read in with `editset`. Try for the rule set in the `peoplerules.txt` file. Also, with a larger rule set, the `summary` and `plot` functions for the `violatedEdits` object become useful.**
```{r}
E <- editfile("data/peoplerules.txt")
ve <- violatedEdits(E, people)
summary(ve)
plot(ve)
```

**Obviously, fixing errors requires subject matter expertise, but the `editrules` package does have facilities for finding potential fixes. The logic goes as follows: Fixing a variable so that an error is eliminated may cause new errors. Also, it makes sense to chose the edits necessary to make a record consistent in a way so as to minimise the total number of edits. This is a minimisation problem and can be automated. Further add-ons for the methods allow finer control by adding confidence weights. A branch-and-bound approach instead of the MIP solver used here is more slow but also allows for more control.**
```{r}
le <- localizeErrors(E, people, method = "mip")
le$adapt
```
**The `le` object contains metadata and a logical array which indicates the minimal set of edits needed to make data follow the rules in `E` (note that `E` is missing at least one rule). The corrections are not part of the `editrules` package.**

---

### 4 - more on the `editrules` package
**This exercise, again borrowed from de Jonge and van der Loo, follows on the previous but has less hand-holding.**

#### 4.1 Reading in and manual checks
**The `dirty_iris` dataset can be found on the course webpage. Read in the data, and make sure that strings are not converted to factors. Replace special values with `NA` and compute the proportion of complete observations.**

#### 4.2 Validating with rules
**Codify the following rules into an `editfile` object:**

  * **Species should be one of the following values: setosa, versicolor or virginica.**
  * **All measurements should be positive.**
  * **Petal length is at least 2 times petal width.**
  * **Sepal length cannot exceed 30 cm.**
  * **The sepals of an iris are longer than its petals.**

**Find out how often each rule is broken. Summarise and plot the object.**

**Use `violatedEdits` to find the observations with too long petals**

**Use `boxplot` and `boxplot.stats` to find observations with outlier values of sepal length. Set outliers to `NA`**

#### 4.3 Correcting data
**Use the `correctWithRules` function from the `deducorrect` library to replace non-positive values of of Petal.Width with NA. **

**Use the (output of the) `LocalizeErrors` function and the `editset` defined in exercise 5.2 to replace all faulty values with NA**
