---
title: "Practical E"
author: "Anne Vinkel, adapted from 'An Introduction to data cleaning with R' by Edwin de Jonge and  Mark van der Loo "
date: "Statistical Programming with R"
output: html_document
---

### 1 - a bit on factors
**Define the vector**
```{r}
v <- factor(c(1,8.2,3,11))
```
**What happens if you apply the function `as.numeric` to `v`? How can you make `v` into a numeric vector?**

```{r}
as.numeric(v)
```
Since `v` is a factor, the visible values in it are actually labels for the underlying factor values - which are nicely ordered consequtive integers. In order to make `v` numeric with the *right* values, use
```{r}
as.numeric(as.character(v))
```


### 2 - reading in data with base R and `readxl`

**The file `unnamed.xls` contains comma-separated values of ages and heights for a set of imaginary persons. Read in the data. Make sure to name both variables, and make both numeric. Hint: The `parse_number` function in the `readr` package is nice** 

```{r}
unnamed <- readxl::read_excel("data/unnamed.xls", 
                   col_names = c("age", "height"))
unnamed

unnamed$height <- readr::parse_number(unnamed$height)
unnamed
```


<!-- ### 3 - matching dirty data against standard labels -->
<!-- **The file `bands.txt` has a list of bands, but somebody messed up the "country" variable. Match it against the vector `c("Australia", "United States", "United Kingdom", "Germany", "Spain")`. Also make sure the `date` variable is a date.** -->

<!-- We read in the data nicely with the `readr` package: -->
<!-- ```{r} -->
<!-- library(readr) -->
<!-- bands <- read_delim("data/bands.txt", delim = "\t", trim_ws = T, -->
<!--                      col_types = cols( date = col_date(format = "%d%b%Y")) ) -->
<!-- ``` -->

<!-- Now for the label matching -->
<!-- ```{r} -->
<!-- countries <- trimws(bands$country) -->
<!-- codes <- c("Australia", "United States", "United Kingdom", "Germany", "Spain") -->

<!-- ## base R match function gives the clean matches -->
<!-- match(countries, codes) -->

<!-- ## Which values are failing?  -->
<!-- countries[is.na(match(countries, codes))] -->

<!-- ## The amatch function allows for some slack -->
<!-- library(stringdist) -->
<!-- amatch(countries, codes, maxDist=3) -->

<!-- ## Which values are still failing? -->
<!-- countries[is.na(amatch(countries, codes, maxDist=3))] -->

<!-- ## We match the closest fit -->
<!-- D <- stringdistmatrix(countries, codes) -->
<!-- i <- apply(D, 1, which.min) -->
<!-- data.frame(rawtext = countries, coded = codes[i]) -->

<!-- ## This may be the time when we hand-code the rest: -->
<!-- bands$country_clean <- codes[amatch(countries, codes, maxDist=3)] -->
<!-- bands$country_clean[ bands$country == "UK"] <- "United Kingdom" -->
<!-- bands$country_clean[ bands$country == "US"] <- "United States" -->
<!-- ``` -->

### 3a - data cleaning
**Exercise 3 is about removing inconsistencies from data. The exercise comes in two versions: One using base procedures by "hand", and one using the editrules package. This first version is less abstract and does not require looking into a new framework. Feel free to skip it and do  3b instead, if it is too simple - or feel free to skip 3b if it's too involved**

**We will work with the following data - you can find it on the course homepage or paste it into a text file from here:**
```{r, eval=F}
age, agegroup, height, status, yearsmarried
21, adult, 6.0, single, -1
2, child, 3, married, 0
18, adult, 5.7, married, 20
221, elderly, 5, widowed, 2
34, child, -7, married, 3
```

```{r}
people <- read.csv( file = "data/people.txt", stringsAsFactors = F, strip.white = T)
```
**We are going to replace impossible values with `NA`:**
* **Age should not be negative and should be less than 150**
* **Height should be greater than zero**
* **Age should be greater than the number of years married**
* **Children cannot be married or widowed**
* **People under 18 of age should be coded as children, those between 18 and 64 should be coded as adults, and those over age 65 should be coded as elderly**
* **Age should be larger than or equal to yearsmarried + 18**
```{r}
library(dplyr)
## Age should not be negative and should be less than 150
people <- people %>% mutate( age = ifelse( age < 0, NA, age))
people <- people %>% mutate( age = ifelse( age>=150, NA, age))

##Height should be greater than zero
people <- people %>% mutate( height = ifelse( height <=0, NA, height))
```
So far so good. But the rest of the rules involve some matter of choice: If age and years married are inconsistent, which do we correct, for instance? For a rule, we could say that between age, age group and status, if two of them agree and the third doesn't, we fix the third.
```{r}
people <- people %>% mutate( agegroup = ifelse(  agegroup ==  "child" & 
                                                 status %in% c("married", "widowed") &
                                                 age >= yearsmarried + 18,
                                                 ifelse(age<=65, "adult", "elderly"),
                                                 agegroup
                                  )) # Here we can actually deduce the value from "age"

people <- people %>% mutate( status = ifelse( agegroup == "child" & 
                                              status %in%  c("married", "widowed") &
                                              age < 18,
                                              "single",
                                              status
                                              )) # Here we can also deduce the value

people <- people %>% mutate(age = ifelse( age<18 & agegroup != "child" & status != "single",
                                          NA,
                                          age
                                          )) # Here we set to missing

# When age and yearsmarried do not agree otherwise, we shall believe in age
people <- people %>% mutate( yearsmarried = ifelse( age >= yearsmarried + 18 | 
                                                    is.na(age) | is.na(yearsmarried) | 
                                                    status == "single"
                                                    , yearsmarried, NA ))

people
```


### 3b - a walkthrough for the `editrules` package 

**This exercise is adapted from an example in De Jonge and van der Loo and walks you through applying functions in the `editrules` package to make a dataset ready for imputation. The `editrules` package allows allows us to define rules which each record of a data set must obey. Editrules can check which rules fail for each record, and find the minimal set of variables to adapt so that all rules can be folowed.** 

**We will work with the following data - you can find it on the course homepage or paste it into a text file from here:**
```{r, eval=F}
age, agegroup, height, status, yearsmarried
21, adult, 6.0, single, -1
2, child, 3, married, 0
18, adult, 5.7, married, 20
221, elderly, 5, widowed, 2
34, child, -7, married, 3
```

```{r}
people <- read.csv( file = "data/people.txt", stringsAsFactors = F, strip.white = T)
```
**For a cleaner workflow, the rules that these data must obey are treated as a separate object, that can then be stored and reapplied (instead, say, of treating these rules and checks as lines of code in a syntax file). We define an edit set and confront the data with these rules:**
```{r, warning=F, message=F}
library(editrules)
E <- editset( c( "age >=0" , "age <= 150" ) )
E
violatedEdits( E, people)
```
**A realistic set of edit rules will be rather larger. It may be defined in a text file and read in with `editset`. Try for the rule set in the `peoplerules.txt` file. Also, with a larger rule set, the `summary` and `plot` functions for the `violatedEdits` object become useful.**
```{r}
E <- editfile("data/peoplerules.txt")
ve <- violatedEdits(E, people)
summary(ve)
plot(ve)
```

**Obviously, fixing errors requires subject matter expertise, but the `editrules` package does have facilities for finding potential fixes. The logic goes as follows: Fixing a variable so that an error is eliminated may cause new errors. Also, it makes sense to chose the edits necessary to make a record consistent in a way so as to minimise the total number of edits. This is a minimisation problem and can be automated. Further add-ons for the methods allow finer control by adding confidence weights. A branch-and-bound approach instead of the MIP solver used here is more slow but also allows for more control.**
```{r}
le <- localizeErrors(E, people, method = "mip")
le$adapt
```
**The `le` object contains metadata and a logical array which indicates the minimal set of edits needed to make data follow the rules in `E` (note that `E` is missing at least one rule). The corrections are not part of the `editrules` package.**


### 4 - more on the `editrules` package
**This exercise, again borrowed from de Jonge and van der Loo, follows on the previous but has less hand-holding. If you want, you can do the exercise manually, without using the `editrules` package - as in exercise 3a**

#### 4.1 Reading in and manual checks
**The `dirty_iris` dataset can be found on the course webpage. Read in the data, and make sure that strings are not converted to factors. Replace special values with `NA` and compute the proportion of complete observations.**
```{r}
dirty.iris <- read.csv( file = "data/dirty_iris.txt", 
                        stringsAsFactors = F)
summary(dirty.iris)

## There are Inf values in Petal.Width
dirty.iris$Petal.Width[ !is.finite(dirty.iris$Petal.Width) ] <- NA

## The complete.cases function checks each record for missing values
prop.table( table( complete.cases(dirty.iris) ))
```
#### 4.2 Validating with rules
**Codify the following rules into an `editfile` object:**

  * **Species should be one of the following values: setosa, versicolor or virginica.**
  * **All measurements should be positive.**
  * **Petal length is at least 2 times petal width.**
  * **Sepal length cannot exceed 30 cm.**
  * **The sepals of an iris are longer than its petals.**

```{r}
iris.rules <- editset(
  c(
    "Species %in% c('setosa', 'versicolor', 'virginica')",
    "Sepal.Length > 0",
    "Sepal.Width  > 0",
    "Petal.Length > 0",
    "Petal.Width  > 0",
    "Petal.Length >= 2*Petal.Width",
    "Sepal.Length <=30",
    "Sepal.Length >= Petal.Length"
  )
)
```
**Find out how often each rule is broken. Summarise and plot the object.**
```{r}
violated.rules.iris <- violatedEdits( iris.rules, dirty.iris)
summary(violated.rules.iris)
plot(violated.rules.iris)
```

**Use `violatedEdits` to find the observations with too long petals**
Which column of the matrix returned by the function corresponds to that rule? We first look at the rules in the rules object
```{r}
iris.rules
```
The rule is "num7". It's easier to pick  out this rule from `violated.rules.iris` if we make it into a tibble
```{r}
library(dplyr)
violated.rules.iris %>% as_tibble() %>%
  filter(num7 == T)

## Find the observations corresponding to this:

## First we find the indices of these observations

indices <- violated.rules.iris %>% 
    as_tibble() %>%
    mutate( id = row_number()) %>%           
    filter(num7 == T)

dirty.iris[indices$id,]
```

**Use `boxplot` and `boxplot.stats` to find observations with outlier values of sepal length. Set outliers to `NA`**
```{r}
## boxplot does a box plot
boxplot(dirty.iris$Sepal.Length)
## boxplot.stats does statistics for the boxplot
bps <- boxplot.stats(dirty.iris$Sepal.Length)
bps

dirty.iris <- dirty.iris %>% 
  mutate( Sepal.Length = if_else(Sepal.Length %in% bps$out, NA_real_, Sepal.Length)) 
```

#### 4.3 Correcting data
**Use the `correctWithRules` function from the `deducorrect` library to replace non-positive values of of Petal.Width with NA. **
```{r, message=F}
library(deducorrect)

# We make a set of correction rules
iris.cr <- correctionRules(expression(
   if ( is.na(Petal.Length) | Petal.Length <= 0 ) Petal.Length <- NA
))

# We use the correction rules on the data
iris.corrections <- correctWithRules( iris.cr , dirty.iris)

iris.corrections$corrections

## It's not very good practise, but we'll replace the data with the corrected data
dirty.iris <- iris.corrections$corrected
```
**Use the (output of the) `LocalizeErrors` function and the `editset` defined in exercise 5.2 to replace all faulty values with NA**
```{r}
iris.loc.err <- localizeErrors( iris.rules , dirty.iris)
plot(iris.loc.err)

# DEGENERATE solutions are those with more than one possible edit. We look at them
table(iris.loc.err$status$degeneracy)
dirty.iris[iris.loc.err$status$degeneracy>1,]
iris.loc.err$adapt[iris.loc.err$status$degeneracy>1,]

# We set all values deemed faulty to NA
dirty.iris[iris.loc.err$adapt] <- NA

summary(violatedEdits( iris.rules, dirty.iris))
```

