---
title: "Statistical Inference"
author: "Jolien Cremers (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    css: css/style.css
    logo: css/logo_en.png
    transition: 0
    self_contained: TRUE   
---

# Statistical Inference

## This Lecture:

- Plotting distributions
- Comparing two groups
- Checking assuptions
- Non-parametric tests

## I use the following package in this lecture
```{r message=FALSE}
library(magrittr) #for pipes
library(MASS)     #for the bacteria data
library(Ecdat)    #for the Benefits data
```





# Plotting distributions
## Plotting curves

Base R function `curve()` to plot the pdf and cdf of a normal distribution:

```{r fig.height=3}
par(mfrow=c(1,2), bg=NA)
curve(dnorm(x,0,1),xlim=c(-3,3),ylab="f(x)",main='') 
curve(pnorm(x,0,1),xlim=c(-3,3),ylab="F(X < x)",main='')
```

## 
or exponential distribution:

```{r echo = FALSE, fig.height=4}
par(mfrow=c(1,2), bg = NA)
curve(dexp(x,2),from=0, to=5,xlab="t",ylab="f(t)",col="red",main='') 
curve(dexp(x,.5),from=0, to=5,main='',col="blue",add=TRUE) 
text(.8,1.8,expression(paste(lambda," = 2")),col="red")
text(4,0.25,expression(paste(lambda," = 0.5")),col="blue")
curve(pexp(x,2),from=0, to=5,xlab="t",ylab="F(T < t)",main='',col="red")
curve(pexp(x,.5),from=0, to=5,main='',col="blue",add=TRUE)
text(1.5,.8,expression(paste(lambda," = 2")),col="red")
text(2.3,0.5,expression(paste(lambda," = 0.5")),col="blue")
```

## Sampling from distributions:

In R we can use:

- `rnorm()` to obtain samples from a normal distribution
- `dnorm()` to sample from a normal density function
- `pnorm()` to sample from a normal distribution function
- `qnorm()` to sample from a normal quantile function

This works for other distributions as well, e.g. `rexp()`, `dexp()`, `pexp()` and `qexp()`.

## Setting a seed

```{r}
set.seed (101)
rnorm(5, 2, 5)
rnorm(5, 2, 5)

set.seed(101)
rnorm(5, 2, 5)
```


## Histograms

```{r echo=FALSE, fig.height=4}
set.seed(101)
hist(rnorm(1000, 3, 0.5), ylab = "P(Y=y)", xlab = "y", ylim = c(0,1), prob = TRUE)
set.seed(101)
lines(density(rnorm(1000, 3, 0.5))) 

```

## Code for previous plot

```{r, eval = FALSE, fig.height=4}

hist(rnorm(1000, 3, 0.5),
     ylab = "P(Y=y)", xlab = "y",
     ylim = c(0,1), prob = TRUE)
lines(density(rnorm(1000, 3, 0.5))) 

```


## Barplots

Barplots for a poisson distribution:

```{r echo=FALSE, fig.height=4}
par(mfrow = c(1, 3), bg = NA)
y <- factor(0:5)
x <- 0:5
barplot(dpois(x, .5), ylim = c(0, .6), names.arg = y, 
        ylab = "P(Y=y)", xlab = "y")
text(5, .5, expression(paste(lambda, " = 0.5")))
barplot(dpois(x, 1), 
        ylim = c(0, .6), 
        names.arg = y, 
        ylab = "P(Y=y)", 
        xlab = "y")
text(5, .5, expression(paste(lambda, " = 1")))
barplot(dpois(x, 2), ylim = c(0, .6), names.arg = y, 
        ylab = "P(Y = y)",xlab="y")
text(5,.5,expression(paste(lambda," = 2")))
```


## Code for the previous plot  {.smaller}
```{r, eval = FALSE, fig.height=4}
par(mfrow = c(1, 3), bg = NA)
y <- factor(0:5)
x <- 0:5
barplot(dpois(x, .5), ylim = c(0, .6), names.arg = y, 
        ylab = "P(Y=y)", xlab = "y")
text(5, .5, expression(paste(lambda, " = 0.5")))
barplot(dpois(x, 1), 
        ylim = c(0, .6), 
        names.arg = y, 
        ylab = "P(Y=y)", 
        xlab = "y")
text(5, .5, expression(paste(lambda, " = 1")))
barplot(dpois(x, 2), ylim = c(0, .6), names.arg = y, 
        ylab = "P(Y = y)",xlab="y")
text(5,.5,expression(paste(lambda," = 2")))
```


# Comparing two groups

## Discrete variables

```{r}
bacteria %>% head(n=8)
```

## Discrete variables

Column `y`: 

* `n`: Disease absent. 
* `y`: Disease present.

Column `ap`: 

* `a`: Active medicine. 
* `p`: Placebo.


## Chi-square test
```{r}
x2.table <- bacteria %$% table(ap, y) #do not need data argument
x2.table

prop.table(table(bacteria$ap, bacteria$y), margin = 1)
```

## Performing a chi-square test
```{r}
bacteria %$% chisq.test(ap, y)
```

## Or, equivalently:
```{r}
bacteria %$% table(ap, y) %>% chisq.test
```

## Expected cell counts

```{r}
bacteria %$% chisq.test(ap, y)$expected
```

## Visualizing this

```{r, dev.args = list(bg = 'transparent')}
bacteria %$% table(y, ap) %>% plot
```

## Visualizing this {.smaller}

```{r fig.height=4.5, dev.args = list(bg = 'transparent')}
barplot(table(bacteria$y, bacteria$ap), 
        beside = TRUE, 
        legend = c("No infection", "Infection"), 
        col = c("blue", "orange"),
        args.legend = list(x=4.5, y=80))
```

## Continuous variables

```{r}
head(Benefits, n = 5)
```

##
Unemployment of blue collar workers:


Column `stateur`: 

* state unemployment rate (in %)

Column `sex`: 

* `male`
* `female`

##

```{r}
hist(Benefits$stateur, xlim=c(2,18),
     main = "State Unemployment Rates in 1972",
     xlab = "State Unemployment Rate (in %)")
```


## Two-sided t-test

```{r}
Benefits %$% t.test(stateur, mu = 5, alternative = "two.sided")
```

## One-sided t-test

```{r}
Benefits %$% t.test(stateur, mu = 5, alternative = "greater")
```

## Two-sample t-test

```{r}
Benefits %$% t.test(stateur~sex)
```

## ANOVA
```{r}
fit <- Benefits %$% aov(stateur~sex)
summary(fit)
```


# Some Assumptions

## Small expected cell frequencies {.smaller}

With small expected cell frequencies (roughly if any are below 5), we should use Fishers exact test.

```{r}
short <- bacteria %>%
  subset(week <= 2) %$%
  table(y, ap)
short
```


## Fisher's exact test

```{r}
short %>% chisq.test
```


## Fisher's exact test

```{r warning=TRUE}
#fisher.test(table(short$y, short$ap))
short %>%
  fisher.test
```

## Assumptions in ANOVA {.smaller}

```{r}
fit <- Benefits %$% aov(stateur~sex)
plot(fit, 2)
```





# Non-parametric testing

## Mann-Whitney / Wilcoxon rank-sum test {.smaller}

Check whether two population distributions (independent samples) are identical without assuming normality.

```{r}
Benefits %$% wilcox.test(stateur~sex, paired = FALSE) 
```

set `paired = TRUE` for Wilcoxon signed-rank test for dependent samples.

## Kruskal-wallis test

Check whether a collection of population distributions (independent samples) are identical without assuming normality.

```{r}
Benefits %$% kruskal.test(stateur~sex)
```


## Some other non-parametric tests in `R`

- `friedman.test`: Friedman rank sum test
- `Rfit` package for nonparametric regression
- `cor(x, y, method = "kendall")` for Kendalls tau-b correlation
- `cor(x, y, method = "spearman")` for Spearman correlation
- `survival` package for semi- and nonparametric methods (Kaplan-Meier/Cox regression) in survival analysis



