---
title: "Practical B"
author: "Jolien Cremers (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

We use the following packages:
```{r}
library(ggplot2)
library(dplyr)
library(magrittr)
library(MASS)
```

---

#### Exercises

---

Again, just like last time it is wise to start with fixing the random seed.

```{r}
set.seed(123)
```

---

1. **Generate two random samples of 10 numbers from a normal distribution with the below specifications. Test the null hypothesis that the population mean is 0.**

- $\mu = 0$ and $\sigma = 2$
- $\mu = 1.5$ and $\sigma = 2$

```{r eval=printAnswers, echo=printAnswers}
x <- rnorm(10, mean = 0, sd = 2)
t.test(x)

x <- rnorm(10, 1.5, 2)
t.test(x)
```

---

2. **Write a function that generates a random sample of `n` numbers from a normal distribution with a user defined mean (i.e. a mean that you can choose when running the function) and standard deviation 1, and returns the `p.value` for the test that the mean is 0.**

```{r eval=printAnswers, echo=printAnswers}
p.value.t <- function (n, mu) {
  x <- rnorm(n, mu, 1)
  t.test(x)$p.value
}

p.value.t(n = 30, mu = 3)
```

---


3. **Use the function of Exercise 2 to generate 50 $p$-values with $n=10,\mu=0$, and make a `qqplot` to compare distribution of the $p$-values with a uniform $[0,1]$ variable.**

```{r eval=printAnswers, echo=printAnswers}
y <- numeric(50)
for (i in 1:50) {
  y[i] <- p.value.t(n = 10, mu = 0)
}

qqplot(x=qunif(ppoints(50)), y)
```

The p-values follow a uniform distribution. 

---

In a study that examined the use of acupuncture to treat migraine headaches, consenting patients on a waiting list for treatment for migraine were randomly assigned in a 2:1:1 ratio to acupuncture treatment, a "sham" acupuncture treatment in which needles were inserted at non-acupuncture points, and waiting-list patients whose only treatment was self-administered (Linde et al., 2005). The "sham" acupuncture treatment was described to trial participants as an acupuncture treatment that did not follow the principles of Chinese medicine. 

---

4. **What is the conclusion when the outcome is classified according to numbers of patients who experienced a greater than 50% reduction in headaches over a four-week period, relative to a pre-randomization baseline?**

Use the following data
```{r}
data <- matrix(c(74, 71, 43, 38, 11, 65), nrow = 2, ncol = 3)
colnames(data) <- c("Acupuncture", "Sham", "Waiting list")
rownames(data) <- c("> 50% reduction", "< 50% reduction")
data
```

We start with calculating the $X^2$-test:
```{r eval=TRUE, echo=TRUE}

X2test <- 
  data %>%
  chisq.test()

X2test
```

which is extremely significant. We can then calculate the expected cell frequencies
```{r}
X2test$expected
```
and the raw residual
```{r}
X2test$observed - X2test$expected
```

as well as the Pearson residual
```{r}
X2test$residuals
```
to infer the difference in observed and expected cell frequencies. Patients on the waiting list experience `> 50% reduction` much less than we would expect under independence of treatment and outcome.

---

5. **Patients who received the acupuncture and sham acupuncture treatments were asked to guess their treatment at the end of their trial. What would you conclude from this data?**

```{r}
data <- matrix(c(82, 17, 30, 30, 26, 16), nrow = 3, ncol = 2)
colnames(data) <- c("Acupuncture", "Sham")
rownames(data) <- c("Chinese", "Other", "Don't know")
data
```

We again start with calculating the $X^2$-test:
```{r eval=TRUE, echo=TRUE}

X2test <- 
  data %>%
  chisq.test()

X2test
```

which is very significant. We can then calculate the expected cell frequencies
```{r}
X2test$expected
```
and the raw residual
```{r}
X2test$observed - X2test$expected
```

as well as the Pearson residual
```{r}
X2test$residuals
```

We find that people who are receiving *true* `Acupuncture` are more inclined to believe that they receive `Chinese` acupuncture than we would expect under independence, while people wo received `Sham` acupuncture are more inclined to believe that they receive `Other` type of acupuncture. `Don't know` is more or less similarly distributed over the observed and expected frequencies.

---

6. **Write a function that chooses automatically whether to do the `chisq.test()` or the `fisher.test()`. Create the function such that it:**

- takes two vectors as input
- creates a table from the two vectors
- checks if there is any expected cell frequency that is smaller than 5
- and that then performs and prints the results of the appropriate test.

```{r}
contingencyTest <- function(x, y) {
  
  # Make a table out of the variables.
  tab <- table(x, y)
  
  #expected frequencies
  exp.freq <- (colSums(tab) %*% t(rowSums(tab))) / sum(tab)
  
  # Choose the correct test. 
  if (any(exp.freq < 5)) {
    results <- fisher.test(x, y)
    return(results)
  } else {
    results <- chisq.test(x, y)
    return(results)
  }
}
```

A more efficient function that tests whether the expected frequencies are smaller than a threshold $t$ is:

```{r}
contingencyTest2 <- function(x, y, t = 5){
  test <- suppressWarnings(chisq.test(x, y))
  if (any(test$expected < t)){
    test <- fisher.test(x, y)
  }
  return(test)
}
```

The `suppressWarnings()` function is used to surpress printing of the warnings that function chi-square puts out to the console when too low expected cell-frequencies are encountered. Since we decide to use Fisher's exact test in those situations anyway, it is redundant to print the message. 

---

7. **Test the function with the dataset `bacteria` (from `MASS`) by testing independence between compliance (`hilo`) and the presence or absence of disease (`y`).**
  
```{r} 
bacteria %$%
  contingencyTest(ap, hilo)
bacteria %$%
  contingencyTest2(ap, hilo)
```

---

8. **Does your function work differently if we only put in the first 25 rows of the `bacteria` dataset?**
```{r}
bacteria[1:25, ] %$%
  contingencyTest(ap, hilo)
bacteria[1:25, ] %$%
  contingencyTest2(ap, hilo)
```
Yes, it performs differently: expected cell frequencies are smaller than 5 for this subset. 
 
---

End of practical. 
