---
title: "Practical B"
author: "Jolien Cremers (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

---

#### Exercises

---

Again, just like last time it is wise to start with fixing the random seed.

```{r}
set.seed(123)
```

---

1. **Generate two random samples of 10 numbers from a normal distribution with the below specifications. Test the null hypothesis that the population mean is 0.**

- $\mu = 0$ and $\sigma = 2$
- $\mu = 1.5$ and $\sigma = 2$

---

2. **Write a function that generates a random sample of `n` numbers from a normal distribution with a user defined mean (i.e. a mean that you can choose when running the function) and standard deviation 1, and returns the `p.value` for the test that the mean is 0.**

---

3. **Use the function of Exercise 3 to generate 50 $p$-values with $n=10,\mu=0$, and make a `qqplot` to compare distribution of the $p$-values with a uniform $[0,1]$ variable.**

---

In a study that examined the use of acupuncture to treat migraine headaches, consenting patients on a waiting list for treatment for migraine were randomly assigned in a 2:1:1 ratio to acupuncture treatment, a "sham" acupuncture treatment in which needles were inserted at non-acupuncture points, and waiting-list patients whose only treatment was self-administered (Linde et al., 2005). The "sham" acupuncture treatment was described to trial participants as an acupuncture treatment that did not follow the principles of Chinese medicine. 

---

4. **What is the conclusion when the outcome is classified according to numbers of patients who experienced a greater than 50% reduction in headaches over a four-week period, relative to a pre-randomization baseline?**

Use the following data
```{r}
data <- matrix(c(74, 71, 43, 38, 11, 65), nrow = 2, ncol = 3)
colnames(data) <- c("Acupuncture", "Sham", "Waiting list")
rownames(data) <- c("> 50% reduction", "< 50% reduction")
data
```

---

5. **Patients who received the acupuncture and sham acupuncture treatments were asked to guess their treatment at the end of their trial. What would you conclude from this data?**

```{r}
data <- matrix(c(82, 17, 30, 30, 26, 16), nrow = 3, ncol = 2)
colnames(data) <- c("Acupuncture", "Sham")
rownames(data) <- c("Chinese", "Other", "Don't know")
data
```

---

6. **Write a function that chooses automatically whether to do the `chisq.test()` or the `fisher.test()`. Create the function such that it:**

- takes two vectors as input
- creates a table from the two vectors
- checks if there is any expected cell frequency that is smaller than 5
- and that then performs and prints the results of the appropriate test.

---

7. **Test the function with the dataset `bacteria` (from `MASS`) by testing independence between compliance (`hilo`) and the presence or absence of disease (`y`).**
  
---

8. **Does your function work differently if we only put in the first 25 rows of the `bacteria` dataset?**

---

End of practical. 
