---
title: "Linear Models"
author: "Jolien Cremers (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: 
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
    font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

---



# The linear model
## This Lecture:

- Linear models in R
- Checking assuptions
- Model fit & comparison
- Predictions

## I use the following packages in this lecture

```{r message = FALSE}

library(magrittr) #for pipes
library(mice)     #for the boys data
library(Ecdat)    #for the Benefits data
library(ggplot2)

```

## The linear model

The function `lm()` is a base function in `R` and allows you to pose a variety of linear models. 
```{r}
args(lm)
```

If we want to know what these arguments do we can ask R:

```{r, eval=FALSE}
?lm
```

This will open a help page on the `lm()` function.

## Continuous predictors {.smaller}
To obtain a linear model with a main effect for `tenure`, we formulate $stateur\sim tenure$ 
```{r warning=FALSE, message = FALSE}
fit <- Benefits %$%
  lm(stateur~tenure)
fit
```

## Continuous predictors: more detail {.smaller}
```{r}
fit %>% summary
```

## Continuous predictors
To obtain a linear model with just main effects for `tenure` and `age`, we formulate $stateur\sim tenure + age$ 
```{r}
require(mice)
fit <- Benefits %$%
  lm(stateur ~ tenure + age)
fit
```

## Continuous predictors: more detail {.smaller}
```{r}
fit %>% summary
```

## Continuous predictors: interaction effects
To obtain a linear model with an interaction effect for `tenure` and `age`, we formulate $stateur\sim tenure * age$ 
```{r}
fit <- Benefits %$% lm(stateur ~ tenure * age)
```

Equivalent
```{r eval = FALSE}
fit <- Benefits %$% lm(stateur ~ tenure + age + tenure:age)
```

## Continuous predictors: interaction effects
```{r}
fit
```

##  and with more detail {.smaller}
```{r}
fit %>% summary
```

## Categorical predictors in the linear model
If a categorical variable is entered into function `lm()`, it is automatically converted to a dummy set in `R`. The first level is always taken as the reference category. If we want another reference category we can use the function `relevel()` to change this.

```{r}
fit_c <- Benefits %$% lm(stateur ~ joblost)
```

##  and again with more detail {.smaller}
```{r}
fit_c %>% summary
```

## Components of the linear model {.smaller}
```{r}
names(fit_c) # the names of the list with output objects
fit_c$coef  # show the estimated coefficients
```

```{r eval = FALSE}
coef(fit_c) # alternative
```





# Checking assumptions
## Linearity {.smaller}
```{r eval = FALSE, message = FALSE}
Benefits %>%
  ggplot(aes(tenure, stateur)) + 
  geom_point() + 
  geom_smooth(method = "loess", col = "blue") + 
  geom_smooth(method = "lm", col = "orange")
```

## Linearity {.smaller}
```{r echo=FALSE, message = FALSE}
Benefits %>%
  ggplot(aes(tenure, stateur)) + 
  geom_point() + 
  geom_smooth(method = "loess", col = "blue") + 
  geom_smooth(method = "lm", col = "orange")
```

The loess curve suggests slight non-linearity

## Adding a squared term
```{r message=FALSE, warning = FALSE}
Benefits %$%
  lm(stateur ~ tenure + I(tenure^2)) %>%
  summary()
```

## Constant error variance? {.smaller}
```{r, fig.height=4, dev.args = list(bg = 'transparent')}
par(mfrow = c(1, 2))
fit %>%
  plot(which = c(1, 3), cex = .6)
```

## No constant error variance! {.smaller}
```{r, fig.height=4, dev.args = list(bg = 'transparent')}
par(mfrow = c(1, 2))
boys %$%
  lm(bmi ~ age) %>%
  plot(which = c(1, 3), cex = .6)
```

## Normality of errors {.smaller}
```{r, fig.height=3.5, dev.args = list(bg = 'transparent')}
fit %>%
  plot(which = 2, cex = .6)
```

The QQplot shows divergence from normality at the tails

## Histograms
```{r, echo = FALSE}
par(mfrow = c(1,2))
hist(Benefits$age, main = "", xlab = "Age")
hist(Benefits$tenure, main = " ", xlab = "Tenure")
```

<!-- ## Outliers and influential cases {.smaller} -->

<!-- Leverage: see the fit line as a lever.  -->

<!--   - some points pull/push harder; they have more leverage -->

<!-- Standardized residuals: -->

<!--   - The values that have more leverage tend to be closer to the line -->
<!--   - The line is fit so as to be closer to them -->
<!--   - The residual standard deviation can differ at different points on $X$ - even if the error standard deviation is constant.  -->
<!--   - Therefore we standardize the residuals so that they have constant variance (assuming homoscedasticity).  -->

<!-- Cook's distance: how far the predicted values would move if your model were fit without the data point in question.  -->

<!--   - it is a function of the leverage and standardized residual  associated with each data point -->

<!-- ## Outliers and influential cases -->
<!-- Outliers are cases with large $e_z$  (standardized residuals). -->

<!-- If the model is ***correct***  we expect: -->

<!--   -  5\% of $|e_z|>1.96$  -->
<!--   -  1\% of $|e_z|>2.58$ -->
<!--   -  0\% of $|e_z|>3.3$  -->

<!-- Influential cases are cases with large influence on parameter estimates -->

<!--   -  cases with Cook's Distance $> 1$, or  -->
<!--   -  cases with Cook's Distance much larger than the rest -->

## Outliers and influential cases {.smaller}
```{r, fig.height= 3, dev.args = list(bg = 'transparent')}
par(mfrow = c(1, 2), cex = .6)
fit %>% plot(which = c(4, 5))
```

There are quite some cases with $|e_z|>2$ (right plot). There are no cases with Cook's Distance $>1$, but some cases stand out.

```{r echo = FALSE}
set.seed(20)

pred1 = rnorm(20, mean=20, sd=3)
outcome1 = 5 + .5*pred1 + rnorm(20)

pred2 = c(pred1, 30);        outcome2 = c(outcome1, 20.8)
pred3 = c(pred1, 19.44);     outcome3 = c(outcome1, 20.8)
pred4 = c(pred1, 30);        outcome4 = c(outcome1, 10)
```

## Fine
```{r echo=FALSE}
par(mfrow = c(1,2))
plot(outcome1 ~ pred1, ylim = c(9, 25), xlim = c(10, 30))
abline(lm(outcome1 ~ pred1))
plot(lm(outcome1 ~ pred1), which = 5)
```

## High leverage, low residual
```{r echo=FALSE}
par(mfrow = c(1,2))
plot(outcome2 ~ pred2, ylim = c(9, 25), xlim = c(10, 30))
points(30, 20.8, col = "red")
abline(lm(outcome2 ~ pred2))
plot(lm(outcome2 ~ pred2), which = 5)
```

## Low leverage, high residual
```{r echo=FALSE}
par(mfrow = c(1,2))
plot(outcome3 ~ pred3, ylim = c(9, 25), xlim = c(10, 30))
abline(lm(outcome3 ~ pred3))
points(19.44, 20.8, col = "red")
plot(lm(outcome3 ~ pred3), which = 5)
```

## High leverage, high residual
```{r echo=FALSE}
par(mfrow = c(1,2))
plot(outcome4 ~ pred4, ylim = c(9, 25), xlim = c(10, 30))
points(30, 10, col = "red")
abline(lm(outcome4 ~ pred4))
plot(lm(outcome4 ~ pred4), which = 5)
```





# Model fit & comparison
## R-squared

```{r }
summary(fit)$r.squared
```

```{r }
summary(fit)$adj.r.squared
```


## AIC and BIC
Akaike's *An Information Criterion* 
```{r }
fit %>% 
  AIC()
```

and *Bayesian Information Criterion*
```{r }
fit %>%
  BIC()
```

## Influence of cases
DfBeta calculates the change in coefficients depicted as deviation in SE's.
```{r }
fit %>%
  dfbeta() %>%
  head(n = 7)
```





# Prediction

## Fitted values:
Let's use the simpler `anscombe` data example
```{r }
fit <- anscombe %$% lm(y1 ~ x1)
y_hat <- 
  fit %>%
  fitted.values()
```
The residual is then calculated as
```{r }
y_hat - anscombe$y1
```

## Predict new values
If we introduce new values for the predictor `x1`, we can generate predicted values from the model
```{r , warning=FALSE}
new.x1 <- data.frame(x1 = 1:20)
predict.lm(fit, newdata = new.x1)
```

## Predictions are draws from the regression line
```{r}
pred <- predict.lm(fit, newdata = new.x1)
lm(pred ~ new.x1$x1)$coefficients
fit$coefficients
```

## Prediction intervals
```{r warning=FALSE}
predict(fit, interval = "prediction")
```

## Some other modeling devices in `R`
- `lm()`: linear modeling
- `glm()`: generalized linear modeling
- `gamlss::gamlss`: generalized additive models (for location, scale and shape)
- `lme4::lmer`: linear mixed effect models
- `lme4::glmer`: generalized linear mixed effect models
- `lme4::nlmer`: non-linear mixed effect models
