---
title: "Practical H"
author: "Jolien Cremers (adapted from Gerko Vink)"
date: "Statistical Programming in R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(MASS)
require(DAAG)
require(ggplot2)
```


---

#### Exercises

---

The following packages are required for this practical:
```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
library(ggplot2)
library(DAAG)
```

---

The data sets `elastic1` and `elastic2` from the package `DAAG` were obtained
using the same apparatus, including the same rubber band, as the data frame
`elasticband`. 

1. **Using a different symbol and/or a different color, plot the data
from the two data frames `elastic1` and `elastic2` on the same graph. Do the two
sets of results appear consistent?**

---

2. **For each of the data sets `elastic1` and `elastic2`, determine the regression of
distance on stretch. In each case determine:**

- fitted values and standard errors of fitted values and
- the $R^2$ statistic.

Compare the two sets of results. What is the key difference between the two sets
of data?

---

3. **Study the *residual vs leverage* plots for both models. Hint use `plot()` on the fitted object**

---

4. **Use the robust regression function `rlm()` from the `MASS` package to fit lines to the data in `elastic1` and `elastic2`. Compare the results with those from use of `lm()`:**

- residuals
- regression coefficients, 
- standard errors of coefficients, 
- plots of residuals against fitted values.

---

5. **Use the `elastic2` variable `stretch` to obtain predictions on the model fitted on `elastic1`.**

---

6. **Now make a scatterplot to investigate similarity between plot the predicted values against the observed values for `elastic2`**


---

7. **Write a function that chooses automatically whether to do the `chisq.test()` or the `fisher.test()`. Create the function such that it:**

- takes two vectors as input
- creates a table from the two vectors
- checks if there is any expected cell frequency that is smaller than 5
- and that then performs and prints the results of the appropriate test.

---

8. **Test the function with the dataset `bacteria` (from `MASS`) by testing independence between compliance (`hilo`) and the presence or absence of disease (`y`).**
  
---

9. **Does your function work differently if we only put in the first 25 rows of the `bacteria` dataset?**

---

The mammalsleep dataset is part of `mice`. It contains the Allison and Cicchetti (1976) data for mammalian species. To learn more about this data, type
```{r}
?mammalsleep
```

--- 

10. **Fit and inspect a model where `brw` is modeled from `bw`**

---

11. **Now fit and inspect a model where `brw` is predicted from both `bw` and `species`**

---

12. **Can you find a model that improves the $R^2$ in modeling `brw`?**

---

13. **Inspect the diagnostic plots for the model obtained in `exercise 12`. What issues can you detect?**

---

End of `Practical H`. 
