---
title: "Practical A"
author: "Anne Vinkel (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

---

#### Exercises

1. **Open Rstudio, read in the `mice` package. What data are in the package? (Hint: `?data` will give you the help file for the `data function`). What else is in the package? **
```{r, message=FALSE}
library(mice)
?data
data( package = "mice")
?`mice-package`
```

---

2. **Look at the data frame `ToothGrowth`  which is part of the `datasets` package included with base R. Print the data, look at the help file, do a summary. How would you look at whether higher doses of vitamin C were associated wih longer teeth? **
```{r}
?ToothGrowth

head(ToothGrowth)

summary(ToothGrowth)

table(ToothGrowth$supp, ToothGrowth$dose)

plot( x = ToothGrowth$dose[ ToothGrowth$supp == "VC" ], 
      y = ToothGrowth$len[ ToothGrowth$supp == "VC" ],
      main = "Vitamin C dose and tooth length, a plot",
      xlab="Dose", ylab="Tooth length")
```
The visual inspection here seems pretty convincing to me. But wouldn't a `ggplot` be prettier? 
```{r, message=F}
library(dplyr)
library(ggplot2)
ToothGrowth %>% 
  filter(supp == "VC") %>% 
  ggplot(aes( x = dose, y = len)) +
  geom_point( size = 2 ) 
```

---

3. **Now look at the table `HairEyeColor`. Look at the help file and print the data. Why does it print weird? How do you make a data frame containing only data on female students? Can you compute the number of brown-haired students? Which hair/eye/sex color combination has the fewest students?**
```{r}
?HairEyeColor
class(HairEyeColor)
HairEyeColor
```

`HairEyeColor` is not just a table, but a 3-dimensional array resulting from cross-tabulating observations on hair color, eye color and sex. Sex makes up the third dimension of the table, so we can get the female students like this: 
```{r}
A <- HairEyeColor[,,2]
A
class(A)
A <- as.data.frame(A)
class(A)
```

Alternatively, you can start by making `HairEyeColor` into a `data.frame` and subset it like this:
```{r}
# Coerce to data.frame
A <- as.data.frame(HairEyeColor) 
head(A)
A[A$Sex == "Female",]
```
The latter, rather impenetrable syntax was base R. With `dplyr`, we can use this (rather prettier) syntax:
```{r, message=FALSE}
library(dplyr)

HairEyeColor %>% 
  as_data_frame() %>% 
  filter( Sex == "Female")
```

```{r}
## Number of brown-haired students
sum(HairEyeColor["Brown",,])

## Which group is the smallest?
min(HairEyeColor)

## Base R
which( HairEyeColor == min(HairEyeColor), arr.ind = TRUE )

## dplyr syntax
HairEyeColor %>% as_tibble() %>% filter(n == min(n))
```

--- 

4. **Make a vector containing the numbers 0 to 0.8 by 0.1. Make a vector repeating the numbers 1,2,3 three times. Make a vector containing the letters A to I. Make a data frame out of these three vectors. Name the columns of the data frame - say, "fractions", "numbers" and "letters". Add a column containing the log of the sum of "numbers" and "fractions". Make a histogram of the logs.**
```{r}
a <- seq( from = 0, to = 0.8, by = 0.1)
b <- rep(1:3, 3)
c <- LETTERS[ 1:9 ]

D <- data.frame(fractions = a, numbers = b, letters = c)
D$logs <- log( D$numbers + D$fractions )

hist(D$logs)
```

---

5. **Save a copy of your data frame to the disk. Where is it saved? Where is your working directory? If it isn't where you want it, move it - say, to a new folder designated for this course**
```{r}
save(D, file = "MyData.RDS")
```
We're saving the table as an `.RDS` file. Unless we specify a path elsewhere, it will be saved in our working directory. 
```{r}
getwd()
```
You can change your working directory with the `setwd` function, but for more portable code, we recommend using Rstudio projects instead. Go to the  "File" menu, and click "new project". Place the project file in the folder you want to work in for this project. When you click "open project" and select this project, Rstudio will set the working directory here until you close the project. 


---

End of practical. 
