---
title: "Practical D"
author: "Anne Vinkel (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: html_document
---

We use the following packages:
```{r warning=FALSE, message=FALSE}
library(mice)
library(dplyr)
library(magrittr)
library(DAAG)
```

---

The following table shows numbers of occasions when inhibition (i.e., no flow of current across a membrane) occurred within 120 s, for different concentrations of the protein peptide-C. The outcome `yes` implies that inhibition has occurred.

    conc 0.1 0.5  1 10 20 30 50 70 80 100 150 
    no     7   1 10  9  2  9 13  1  1   4   3 
    yes    0   0  3  4  0  6  7  0  0   1   7

---

1. **Create this data in `R`**
```{r}
data <- data.frame(conc = c(.1, .5, 1, 10, 20, 30, 50, 70, 80, 100, 150),
                   no = c(7, 1, 10, 9, 2, 9, 13, 1, 1, 4, 3),
                   yes = c(0, 0, 3, 4, 0, 6, 7, 0, 0, 1 ,7)) 
data
```


---

2. **Add two new variables (columns) to the data**

- the margin of `no` and `yes` over the rows (i.e. `no` + `yes`)
- the proportion (`yes` over `margin`)
<!-- - the logodds -->

<!-- First, we create a function to calculate the logit (logodds): -->
<!-- ```{r} -->
<!-- logit <- function(p) log(p / (1 - p)) -->
<!-- ``` -->

<!-- Then we add the new columns -->
```{r}
data <- 
  data %>% 
  mutate(margin = no+yes,
         prop = yes / margin)
```

<!-- --- -->

<!-- 3. **Inspect the expanded data. What do you see?** -->
<!-- ```{r} -->
<!-- data -->
<!-- ``` -->

<!-- There are a lot of zero proportions, hence the $-\infty$ in the logit. You can fix this (at least the interpretation of the `logodds`) by adding a constant (usually 0.5) to all cells conform the empirical `logodds` [(see e.g. Cox and Snell 1989)](http://www.amazon.com/Analysis-Edition-Monographs-Statistics-Probability/dp/0412306204).  -->

<!-- --- -->

<!-- 4. **Add a new column where the log odds are calculated as:** -->
<!-- $$\log\text{odds} = \log\left(\frac{\text{yes} + 0.5}{\text{no} + 0.5}\right)$$ -->
<!-- ```{r} -->
<!-- logitCandS <- function(yes, no) log((yes + .5) / (no + .5)) -->
<!-- data <-  -->
<!--   data %>%  -->
<!--   mutate(logitCS = logitCandS(yes, no)) -->
<!-- data -->
<!-- ``` -->
<!-- We can now see that the $-\infty$ proportions are mostly gone -->

---

3. **Fit the logistic model with `margin` as the weights**
```{r}
fit <- 
  data %$%
  glm(prop ~ conc, family=binomial, weights=margin)
```

---

4. **Look at the summary of the fitted object**
```{r}
summary(fit)
```

---

5. **Inspect the plots number 1 and 5 for `fit`**
```{r}
plot(fit, which = c(1, 5))
```

The data set is small, but case `11` stands out in the `Residuals vs. Leverage` plot

---

6. **`conc` is somewhat skewed. Run the model again with a log-transformation for `conc`.**
```{r echo = FALSE}
data$conc %>%
  density() %>%
  plot()
```


To apply this in the model directly:

```{r}
fit.log <- 
  data %$%
  glm(prop ~ I(log(conc)), family=binomial, weights=margin)
```

---

7. **Look at the summary of the fitted object again**
```{r}
summary(fit.log)
```
The logodds now depict the unit increase in `log(conc)`, instead of `conc`.

---

8. **Inspects the plots number 1 and 5 of the fitted object based on `log(conc)`.**
```{r}
plot(fit.log, which = c(1, 5))
```

Outliers are now less of an issue. 

---

9. **Use the `brandsma` data from package `mice` to fit a logistic regression model for `sex` based on `lpo` (Language Post Outcome).**
```{r}
brandsma.subset <- 
  brandsma %>%
  subset(!is.na(sex) & !is.na(lpo), select = c(sex, lpo))
fit <- 
  brandsma.subset %$%
  glm(sex ~ lpo, family=binomial(link='logit'))
fit %>%
  summary()
```

With every unit increase in `lpo`, the logodds of gender increases by `r coef(fit)[2]`. 

---

10. **Obtain confidence intervals for the parameter estimates.**
```{r}
confint(fit)
```

---

11. **Use the model parameters to predict the `sex` variable and compare your predictions to the observed `sex`. **
We can obtain predictions by using function `predict`. The default predictions are on the scale of the linear predictors; the alternative "response" is on the scale of the response variable. Thus for a default binomial model the default predictions are of log-odds (probabilities on logit scale) and type = "response" gives the predicted probabilities. 
To obtain the predicted logodds:
```{r}
pred.logodds <- 
  fit %>%
  predict()
head(pred.logodds)
```
and the predicted probabilities
```{r}
pred.prob <- 
  fit %>%
  predict(type = "response")
head(pred.prob)
```
We can then use the decision boundary `pred.prob > .5` to assign cases to `sex == 1` and the others to `sex == 0`. 
```{r}
pred <- numeric(length(pred.prob))
pred[pred.prob > .5] <- 1
pred <- unlist(pred)
```
To determine how many correct predictions we have, we can use
```{r}
obs <- 
  brandsma %>%
  subset(!is.na(sex) & !is.na(lpo), select = sex)
sum(pred == obs) #total correct
mean(pred == obs) #average correct
```
So we succesfully predict about half. Not that impressive. An alternative way to test our model efficiency is with `CVbinary()`:
```{r}
CVbinary(glm(sex ~ lpo, family=binomial(link='logit'), data = brandsma.subset))
```
---

The lesson here is that a significant parameter has no meaning if the substantive interpretation of the effect is ignored. There is almost no relation, whatsoever, there is just sufficient data to deem the influence of `lpo` on `sex` worthy of significance. 

---

12. 
The `moths` data frame in the `DAAG` package contains data from a study of the effect of habitat on the densities of two species of moth (A and P). The quasipoisson distribution is a reasonable model for these data. Try modeling the variable `A` using this distribution. Is there a better link function than the standard?
```{r}
A.glm <- glm(formula = A ~ log(meters) + factor(habitat), family =
quasipoisson, data = moths)
summary(A.glm)
# Note the huge standard errors

## Consider the square root link as another possibility
A2.glm <- glm(formula = A ~ sqrt(meters) + factor(habitat), family = quasipoisson(link=sqrt), data = moths)
summary(A2.glm)
```

---
13. 
Let's try to specify our own link function: $f(x) = \log(\exp(x)-1)$. Generate data like this: 
```{r}
n <- 1000                       
x <- runif(n)
sh <- 2                        
y <- rgamma(n,scale=log(2+3*x)/sh,shape=sh)
```
Specify the link function and run a `glm` regressing `y` on `x`. **Hint:** Look at the help file for `family`. You need to make an object of type "link-glm", containing the items needed for `glm` to run: a link function, an inverse link, the derivative of the inverse link, and a function that describes the domain of the inverse link. 

```{r}
 ## link
 linkfun <- function(y) log(exp(y)-1)
 ## inverse link
 linkinv <- function(eta)  log(exp(eta)+1)
 ## derivative of invlink wrt eta
 mu.eta <- function(eta) { 1/(exp(-eta) + 1) }
 valideta <- function(eta) TRUE
 link <- "log(exp(y)-1)"
 vv <- structure(list(linkfun = linkfun, linkinv = linkinv,
                 mu.eta = mu.eta, valideta = valideta, 
                 name = link),
                 class = "link-glm")

glm(y~x,family=Gamma(link=vv)) 

```


End of Practical