---
title: "Practical D"
author: "Anne Vinkel (adapted from Gerko Vink)"
date: "Statistical Programming with R"
output: html_document
---

We use the following packages:
```{r warning=FALSE, message=FALSE}
library(mice)
library(dplyr)
library(magrittr)
library(DAAG)
```

---

The following table shows numbers of occasions when inhibition (i.e., no flow of current across a membrane) occurred within 120 s, for different concentrations of the protein peptide-C. The outcome `yes` implies that inhibition has occurred.

    conc 0.1 0.5  1 10 20 30 50 70 80 100 150 
    no     7   1 10  9  2  9 13  1  1   4   3 
    yes    0   0  3  4  0  6  7  0  0   1   7

---

1. **Create this data in `R`**

---

2. **Add two new variables (columns) to the data**

- the margin of `no` and `yes` over the rows (i.e. `no` + `yes`)
- the proportion (`yes` over `margin`)

---

3. **Fit the logistic model with `margin` as the weights**

---

4. **Look at the summary of the fitted object**

---

5. **Inspect the plots number 1 and 5 for `fit`**

---

6. **`conc` is somewhat skewed. Run the model again with a log-transformation for `conc`.**


---

7. **Look at the summary of the fitted object again**

---

8. **Inspects the plots number 1 and 5 of the fitted object based on `log(conc)`.**

---

9. 
The `moths` data frame in the `DAAG` package contains data from a study of the effect of habitat on the densities of two species of moth (A and P). The quasipoisson distribution is a reasonable model for these data. Try modeling the variable `A` using this distribution. Is there a better link function than the standard?

---

10. 
Let's try to specify our own link function: $f(x) = \log(\exp(x)-1)$. Generate data like this: 
```{r}
n <- 1000                       
x <- runif(n)
sh <- 2                        
y <- rgamma(n,scale=log(2+3*x)/sh,shape=sh)
```
Specify the link function and run a `glm` regressing `y` on `x`. **Hint:** Look at the help file for `family`. You need to make an object of type "link-glm", containing the items needed for `glm` to run: a link function, an inverse link, the derivative of the inverse link, and a function that describes the domain of the inverse link. 

---

End of Practical