---
title: "Imputation"
author: "Jolien Cremers"
date: "Statistical Programming with R"
output:
  ioslides_presentation:
    css: css/style.css
    logo: css/logo_en.png
    transition: 0
    self_contained: TRUE   
---

# Imputation

## This Lecture:

- Missing data patterns
- Imputation methods

## I use the following packages in this lecture

```{r message = FALSE}

library(mice) # for visualisation & pmm
library(VIM) # visualisation & several hotdeck imputation methods (fast implementation)
library(simputation) # common interface for several imputation methods
library(magrittr) # for pipe operator
library(naniar) # additional package for practical
library(visdat) # visualisation of datasets
library(UpSetR) # Upset plots
library(ggplot2) # plots

```

# Missing data patterns

## `nhanes` data {.smaller}

```{r}
head(nhanes, n = 10)
```

##
```{r}
summary(nhanes)
```


## Visualise missing data pattern {.smaller}

```{r}

VIM::aggr(nhanes, col=c('blue','red'), numbers = TRUE, sortVars = TRUE,
                  labels = names(nhanes), cex.axis = .8, gap = 3, 
                  ylab = c("Histogram of missing data", "Red is Missing Data"))

```


## 
```{r}
pattern <- mice::md.pattern(nhanes)
```

##
```{r}
pattern
```

##
```{r}
visdat::vis_miss(nhanes)
```

##
```{r}
naniar::gg_miss_upset(nhanes)
```

##
```{r, message = FALSE, warning = FALSE}
naniar::gg_miss_fct(x = nhanes, fct = age)
```


## Both variables observed
```{r}
pairs <- mice::md.pairs(nhanes)

pairs$rr
```

## Row missing, column observed
```{r}
pairs$mr
```

## Both variables missing
```{r}
pairs$rr
```


# Imputation methods

## Specifying models in `simputation`

Common structure for specifying an imputation model:

```{r, eval = FALSE}
impute_<model>(data, formula, [model-specific options])
```

Structure of the model in `formula`:

```{r, eval = FALSE}
IMPUTED ~ MODEL_SPECIFICATION [ | GROUPING ]
```


## Regression imputation

```{r}
imp_1 <- impute_lm(nhanes, bmi ~ age + hyp + chl)
head(imp_1)
```

## Median imputation

```{r}
imp_2 <- simputation::impute_median(imp_1, bmi ~ age)
head(imp_2)
```

##

```{r}

imp_2a <- nhanes %>%
          simputation::impute_lm(bmi ~ age + hyp + chl) %>%
          simputation::impute_median(bmi ~ age)
head(imp_2a)
```

## Decision tree imputation

```{r}
imp_3 <- impute_cart(nhanes, hyp ~ .)
head(imp_3)
```

## Other methods

- `impute_hotdeck()`:  random and sequential hot deck, k-nearest neighbours imputation and predictive mean matching.
- `impute_multivariate()`: imputation based on EM-estimation of multivariate normal parameters, imputation based on iterative Random Forest estimates and stochastic imptuation based on bootstrapped EM-estimatin of multivariate normal parameters.
- `impute_proxy()`: Impute missing values by a constant, by copying another variable computing transformations from other variables.


## Predictive Mean Matching (mice)

Default imputation

```{r}
imp <- mice(nhanes, print = FALSE)
```

Defaults:

  - pmm, predictive mean matching (numeric data)
  - logreg, logistic regression imputation (binary data, factor with 2 levels)
  - polyreg, polytomous regression imputation (unordered cateogrical data, factor > 2 levels)
  - polr, proportional odds model (ordered data, factor > 2 levels)

## Check imputations

```{r}
stripplot(imp, chl, pch = 19, xlab = "Imputation number")
```

## Vary number of imputations

Just 4 imputed datasets instead of the default (5)

```{r}
imp <- mice(nhanes, m = 4, print = F)
```

## Predictor matrix

```{r}
imp$pred
```

1 = column used to impute row

## Initial predictor matrix

```{r}
ini <- mice(nhanes, maxit = 0, print = F)
pred <- ini$pred
pred
```


## Change predictor matrix 

We do not use `hyp` to impute any other variables

```{r}
pred[ ,"hyp"] <- 0
pred
imp <- mice(nhanes, m = 4, init = pred, print = F)
```

## Convergence

```{r}
plot(imp)
```

## Seed values

```{r}

imp <- mice(nhanes, m = 4, init = pred, print = F, seed = 101)

```

## Changing imputation methods

```{r}
imp$meth
str(nhanes)
```

##

```{r}
str(nhanes2)
imp <- mice(nhanes2, m = 4, init = pred, print = F, seed = 101)
imp$meth
```

## Available methods {.smaller}

```{r, warning = FALSE, message = FALSE}
methods(mice)
```

## Extracting imputed data

The 3rd imputed dataset
```{r}
c3 <- complete(imp, 3) 
md.pattern(c3, plot = FALSE)
```

## {.smaller}
All imputed datasets in long format

```{r}
complete(imp, "long")
```


## Pooling analyses {.smaller}

```{r}
fit <- with(imp, lm(bmi ~ chl))
fit
```

##

```{r}
class(fit)
ls(fit)
```

##

```{r}
summary(fit$analyses[[2]])
```

##

```{r}
pool.fit <- pool(fit)
summary(pool.fit)
```

## Other packages for imputation

- `mi`
- `Hmisc`
- `Amelia`
- `mix`

