---
title: "Practical F"
author: "Jolien Cremers"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

----

The purpose of this practical is to use some of R's packages to solve a missing data problem by imputing missing values in a 'real' dataset. This can either be a dataset you've brought to this course yourself or the `selfreport` data from the `mice` package. The 'selfreport' data is described in Section 7.3 of the first edition of the book Flexible Imputation of Missing Data (van Buuren, S. (2012). *Flexible Imputation of Missing Data.* CRC/Chapman & Hall, FL: Boca Raton).

---

Suggested packages:

```{r message = FALSE}

library(mice) # for visualisation & pmm
library(VIM) # visualisation & several hotdeck imputation methods (fast implementation)
library(simputation) # common interface for several imputation methods
library(magrittr) # for pipe operator
library(dplyr)
library(naniar) # additional package for practical
library(visdat) # visualisation of datasets
library(UpSetR) # Upset plots
library(ggplot2) # plots

```

---

#### Selfreport data

---

The selfreport data contains data from two studies, one from the `mgg` study, the *survey dataset*, where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study, the *calibration dataset*, where the height and weight measures are both selfreported and measured (variables `hm` and `wm`). The goal for the self-report data is to obtain correct estimates of the prevalence of obesity (Body Mass Index > 32) in the *survey dataset*. Note that the Body Mass Index (bmi) is computed as follows:

$$\text{bmi} = \frac{\text{weight in kg}}{(\text{height in m})^2} $$

---

1. **Check missing data patterns.**

First, check if there is a description of the `selfreport` data in the `mice` package documentation:

```{r}
?selfreport
```

The selfreport data contains data from two studies, one from the `mgg` study where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study where the measures are both selfreported and measured (variables `hm` and `wm`). 

Then, investigate the missing data patterns

```{r}
head(selfreport, n = 10)
```

```{r}
summary(selfreport)
```

The measures weight and height variables are indeed missing from the `mgg` study

```{r}
VIM::aggr(selfreport, col=c('blue','red'), numbers=TRUE, sortVars=TRUE,
                  labels=names(selfreport), cex.axis=.8, gap=3, 
                  ylab=c("Histogram of missing data", "Red is Missing Data"))
```

Apart from measured height and weight also the measured bmi, the indicators for pregnancy, educational level and ethnicity contain missing values.

---

2. **Check the relation between measured and self reported bmi in the *calibration dataset*.**

There are more individuals that have a lower selfreported BMI than measured BMI.

```{r}

calibration <- selfreport %>% 
  filter(as.numeric(src) == 1) %>% 
  mutate(difference = case_when(br < bm ~ "lower",
                                bm < br ~ "higher",
                                bm == br ~ "same"))

ggplot(calibration, aes(br, bm, color = difference)) +
  geom_point() +
  theme_minimal() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(x = "Selfreported BMI", y = "Measured BMI", fill = "")

```



---

3. **Impute missing data.**

First define a deterministic function to compute bmi:

```{r}
bmi <- function(h, w) {
  return(w/(h/100)^2)
}
```

Define the imputation methods to use (we only impute `wm`, `hm` and `bm`). Make
sure to define the deterministic relation between height, weight and bmi for the
computation (not imputation) of bmi.

```{r}
meth <- c(rep("", 5), "pmm", "pmm", rep("", 6), "~bmi(hm,wm)", "")
```

Get predictor matrix:
```{r, warning = FALSE, message = FALSE}
init <- mice(selfreport, maxit = 0) #get initial predictor matrix
pred <- init$pred #note that pregnancy is not imputed by default because it contains too many missings
pred[, c("src", "id", "pop", "prg", "edu", "etn", "web", "bm", "br")] <- 0 #set variables not used as predictors to 0
```

Impute:
```{r, results = FALSE}
imp <- mice(selfreport, pred = pred, meth = meth, seed = 66573, maxit = 20, m = 10)
```


---

4. **What is the estimated prevalence (selfreported and imputed data) of obesity for different `age` and `sex` in the *survey dataset*?**

Estimated prevalence in the selfreported data

```{r}

reported <- as.tbl(selfreport)

reported %>% 
  filter(as.numeric(src) == 2) %>% 
  mutate(br = bmi(hr, wr)) %>%
  mutate(obese = case_when(br > 32 ~ 1,
                           TRUE ~ 0)) %>% 
  mutate(age_cat = cut(age, c(17, 29, 39, 49, 59, 75))) %>% 
  group_by(sex, age_cat) %>% 
  summarise(N = n(), prevalence = sum(obese)/N*100)

```

Estimated prevalence in the imputed data

```{r}

imputed <- as.tbl(complete(imp, action = "long"))

imputed %>% 
  filter(as.numeric(src) == 2) %>% 
  mutate(obese = case_when(bm > 32 ~ 1,
                           TRUE ~ 0)) %>% 
  mutate(age_cat = cut(age, c(17, 29, 39, 49, 59, 75))) %>% 
  group_by(sex, age_cat) %>% 
  summarise(N = n()/10, prevalence = sum(obese)/(N*10)*100)

```

---

End of practical. 
