---
title: "Practical F"
author: "Jolien Cremers"
date: "Statistical Programming with R"
output: html_document
---

---
title: "Practical F"
author: "Jolien Cremers"
date: "Statistical Programming with R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

---

The purpose of this practical is to use some of R's packages to solve a missing data problem in a 'real' dataset. This can be either the dataset you've brought to this course yourself or the 'selfreport' data from the `mice` package. The 'selfreport' data is used in Section 7.3 of the first edition of the book Flexible Imputation of Missing Data (van Buuren, S. (2012). *Flexible Imputation of Missing Data.* CRC/Chapman & Hall, FL: Boca Raton).

---

Suggested packages:

```{r message = FALSE}

library(mice) # for visualisation & pmm
library(VIM) # visualisation & several hotdeck imputation methods (fast implementation)
library(simputation) # common interface for several imputation methods
library(magrittr) # for pipe operator
library(naniar) # additional package for practical
library(visdat) # visualisation of datasets
library(UpSetR) # Upset plots
library(ggplot2) # plots

```

---

#### Selfreport data

---

The selfreport data contains data from two studies, one from the `mgg` study, the *survey dataset*, where height and weight are selfreported measures (`hr` and `wr`) and one from the `krul` study, the *calibration dataset*, where the height and weight measures are both selfreported and measured (variables `hm` and `wm`). The goal for the self-report data is to obtain correct estimates of the prevalence of obesity (Body Mass Index > 32) in the *survey dataset*. Note that the Body Mass Index (bmi) is computed as follows:

$$\text{bmi} = \frac{\text{weight in kg}}{(\text{height in cm})^2} $$

---

1. **Check missing data patterns.**

---

2. **Check the relation between measured and self reported bmi in the *calibration dataset*.**

---

3. **Impute missing data.**

---

4. **What is the estimated prevalence (selfreported and imputed data) of obesity for different `age` and `sex` in the *survey dataset*?**

---

End of practical. 
